\thetitle

### 2013-06-19 tfpie/tfpie.tex
Some of Sam's edits.

      ;; Posn -> Number
      ;; Distance between this posn and that posn
      (check-expect ((new posn 0 0) . dist (new posn 3 4)) 5)
      (define (dist that)
        (sqrt (+ (sqr (- (this . x) (that . x)))
                 (sqr (- (this . y) (that . y))))))

      ;; -> Number
      ;; Distance of this posn from the origin
      (check-expect ((new posn 0 0) . dist-origin) 0)
      (check-expect ((new posn 3 4) . dist-origin) 5)
      (define (dist-origin)
        (this . dist (new posn 0 0))))
\end{verbatim}

This class definition defines a new class of values which are {\tt
  posn} objects.  Such objects are comprised of two numeric values
and understand the messages {\tt x}, {\tt y}, {\tt dist}, and {\tt
  dist-origin}.  Unit tests have been included with each method
definition, following the principles of the design recipe studied in
the first semester.  In fact, the {\tt check-expect} mechanism works
exactly as it did before.

Methods can be defined to consume any number of arguments, but they
are implicitly parameterized over {\tt this}, the object that was sent
the message.

@@

The structure of this data definition is analogous to the approach of
the previous semester but this example brings to light an important
difference with the functional approach.  The method for computing the
sum of the empty tree is defined in the {\tt leaf} class, while the
method for computing the sum of a node is in the {\tt node} class.
When a tree object is sent the {\tt sum} method, there is no function
with a conditional to determine whether the object is a
leaf---instead, the object itself takes care of computing the sum
based on the {\tt sum} method it contains.  This shift in perspective
is at the core of object-orientation: objects contain their own
behavior and the case analysis previously done in functions is
eliminated.

@@

Programming in the first semester is often oriented around interactive
event-driven video games.  The basic design of a video game involves
defining a data representation for states of the game and functions
for transitioning between states based on events such as clock ticks,
keyboard input, or mouse events.  The design of a game thus involves
the design of data and functions on that data; in other words, the
game involves the design of objects.  We therefore continue in the
second semester with the use of programming video games but supplement
the course with a library for doing so in an object-oriented style.
Figure~\ref{fig:world} gives an example written in both the functional
style and object-oriented style.

The key difference between these two programs is that the functional
program uses the {\tt 2htdp/universe} library, which provides a {\tt
  big-bang} form that consumes the initial state of the world and has
a declarative form of associating event-handler functions.  The
object-oriented program uses an alternative library developed for the
class language: {\tt class/universe}.  It also provides a {\tt
  big-bang} form but it consumes a single argument, the initial state
of the world represented as an object.  Event handlers are just
methods of this object.

The program on the left is the first program of the first semester,
while the one on the right is the first program of the second
semester.  Our approach is able to make the conceptual connection
between functional and object-oriented programming quite clear while
appealing the familiar event-driven interactive programs developed
throughout the year.

The move to object-oriented style immediately and naturally leads to
design principles that are enabled by organizing programs around
objects.  For example, the state-pattern becomes useful almost
immediately.  The programs in figure~\ref{fig:world} animate a rocket
(rendered as a circle in this example) taking off.  An illustrative
follow-up exercise is to animate a rocket that \emph{lands}.  The
natural design is to have two variants for states of the rocket: one
for descending rockets, one for landed rockets.  While in the
functional approach it is easy to use the state-pattern for the
\emph{data} representing a rocket, it is more difficult to have states
of \emph{behavior}.  Of course in the object approach, states of
behavior are just as natural as data.  Therefore it is straightforward
to design programs with easy to observe invariants such as ``a landed
rocket never changes position.''  In the functional approach, even such
simple properties are more involved to establish.

@@

% \nocite{*}
\bibliographystyle{plainnat}
\bibliography{dvh-bibliography,local}
\end{document}

>>>>

      ;; dist : Posn -> Number
      ;; Distance between this posn and that posn
      (check-expect ((new posn 0 0) . dist (new posn 3 4)) 5)
      (define (dist that)
        (sqrt (+ (sqr (- (this . x) (that . x)))
                 (sqr (- (this . y) (that . y))))))

      ;; dist-origin : -> Number
      ;; Distance of this posn from the origin
      (check-expect ((new posn 0 0) . dist-origin) 0)
      (check-expect ((new posn 3 4) . dist-origin) 5)
      (define (dist-origin)
        (this . dist (new posn 0 0))))
\end{verbatim}

This class definition defines a new class of values which are {\tt
  posn} objects.  Such objects are comprised of two numeric values and
understand the messages {\tt x}, {\tt y}, {\tt dist}, and {\tt
  dist-origin}.  Unit tests have been included with each method
definition, following the principles of the design recipe studied in
the first semester.  Although {\tt check-expect} forms can appear
within class definitions, they are lifted to the top-level when a
program is run.

Methods can be defined to consume any number of arguments, but they
are implicitly parameterized over {\tt this}, the object that received
the message.

@@

The structure of this data definition is analogous to the approach of
the previous semester but this example brings to light an important
difference with the functional approach.  The method for computing the
sum of a leaf is defined in the {\tt leaf} class, while the
method for computing the sum of a node is in the {\tt node} class.
When a tree object is sent the {\tt sum} method, there is no function
with a conditional to determine whether the object is a
leaf---instead, the object itself takes care of computing the sum
based on its own {\tt sum} method.  This shift in perspective
is at the core of object-orientation: objects contain their own
behavior and the case analysis previously done in functions is
eliminated.

@@

At Northeastern, Programming in the first semester is often oriented
around interactive event-driven video games.  The basic design of a
video game involves defining a data representation for states of the
game and functions for transitioning between states based on events
such as clock ticks, keyboard input, or mouse events.  The design of a
game thus involves the design of data and operations on that data; in
other words, the game involves the design of objects.  We therefore
continue in the second semester with the use of programming video
games but supplement the course with a library for doing so in an
object-oriented style.  Figure~\ref{fig:world} gives an example
written in both the functional style and object-oriented style.

The key difference between these two programs is that the functional
program uses the {\tt 2htdp/universe} library, which provides a {\tt
  big-bang} form that consumes the initial state of the world and has
a declarative form of associating event-handler functions, while the
object-oriented program uses an alternative library developed for our
class: {\tt class/universe}.  It also provides a {\tt big-bang} form
but it consumes a single argument, the initial state of the world
represented as an object.  Event handlers are simply methods of this
object.

The program on the left is the first program of the first semester,
while the one on the right is the first program of the second
semester.  Our approach is able to make the conceptual connection
between functional and object-oriented programming quite clear while
appealing to the familiar event-driven interactive programs developed
throughout the year.

The move to object-oriented style immediately and naturally leads to
designs that are enabled by organizing programs around objects.  For
example, the state pattern~\cite{samth:GOF} becomes useful almost
immediately.  The programs in figure~\ref{fig:world} animate a rocket
(rendered as a circle in this example) taking off.  An illustrative
follow-up exercise is to animate a rocket that \emph{lands}.  The
natural design is to have two variants for states of the rocket: one
for descending rockets, one for landed rockets.  While in the
functional approach it is easy to use the state-pattern for the
\emph{data} representing a rocket, it is more difficult to have states
of \emph{behavior}.  The typical solution adds conditionals to all of
the event handlers.  In the object-oriented approach, states of
behavior are just as natural as data.  It is therefore straightforward
to design programs with easy to observe invariants such as ``a landed
rocket never changes position.''  In the functional approach, even
such simple properties are more involved to establish.

@@

% \nocite{*}
\bibliographystyle{plainnat}
\bibliography{dvh-bibliography,local,sth-bibliography}
\end{document}
